Finally Tagless, Partially Evaluated

### 2007-07-19 tagless/tagless-final.tex
Done with reinflating the paper for JFP, except for related work

A popular way to define and implement a language is to embed it in
another \citep{reynolds-definitional}.  Embedding means to represent
terms and values of the \emph{object language} as terms and values in the
\emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
specific object languages because it supports rapid prototyping and integration
with the host environment \citep{hudak-building}.
If the metalanguage supports \emph{staging}, then
the embedding can compile object programs to the metalanguage and avoid the
overhead of interpreting them on the fly \citep{WalidICFP02}.  A staged
definitional interpreter is thus a promising way to build a domain\hyp specific
language.

@@

We represent object programs using ordinary functions rather than
data constructors.  These functions comprise the entire interpreter, shown
below.
\ifshort
\begin{code3}[xleftmargin=0pt]
let varZ env    = fst env       let b (bv:bool) env = bv
let varS vp env = vp (snd env)  let app e1 e2 env   = (e1 env) (e2 env)
let lam e env   = fun x -> e (x,env)
\end{code3}
\else
\begin{code}
let varZ env        = fst env
let varS vp env     = vp (snd env)
let b (bv:bool) env = bv
let lam e env       = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code}
\fi
We now represent our sample term $(\fun{x}x)\True$ as
\ifshort
\texttt{let testf1 = app (lam varZ) (b true)}.
\else
\begin{code}
let testf1 = app (lam varZ) (b true)
\end{code}
\fi
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment\ifshort
, |testf1 ()|, obtaining |true|\fi.
\ifshort\else
\begin{code}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code}
\fi
The result has no tags: the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
open term such as
\ifshort
\texttt{app (lam (varS varZ)) (b true)}
\else
|testf3| below
\fi
gives a type error rather than a run-time error.
\ifshort\else
\begin{code}
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code}
\fi
The type error correctly complains
that the initial environment should be a tuple rather than~|()|.
In other words, the term is open.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN) CPS, so to isolate the evaluation
        order of the object language from that of the metalanguage.\ifshort
\footnote{Due to serious lack of space, 
we refer the reader to the accompanying code for this.}\fi
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
MetaOCaml) and standard Haskell. The complete code is available at
\url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.

@@

To reiterate, we do \emph{not} propose any new language
feature or technique. We use features already present in
mainstream functional languages---Hindley-Milner
type system with either an inference\hyp preserving module system or
constructor classes, as realized in ML and Haskell 98---to
solve the problem that was stated in the published record as
unsolved and likely unsolvable in ML or Haskell 98
without extensions. The simplicity of our solution and our
use of only mainstream features makes it more practical to build typed,
embedded domain\hyp specific languages.
\begin{comment}
We may claim some contribution about type-preserving CPS. Check
related work section in \cite{Guillemette-Monier-PLPV}, especially
check the work of Shao on type-preserving CPS in Flint. The PLPV paper
in related work shows other tasks, including closure conversion, which
we may tackle in our approach. We may be able to write a
type-preserving, assured compiler, whose properties and assured by HM.
\end{comment}

@@

\begin{figure*}[t]
\begin{floatrule}
\begin{code}
module type Symantics = sig type ('c, 'dv) repr
  val int : int  -> ('c, int) repr
  val bool: bool -> ('c, bool) repr

@@

  val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val mul : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
  val if_ : ('c,bool) repr -> (unit -> ('c, 'da) repr)
                           -> (unit -> ('c, 'da) repr) -> ('c, 'da) repr
end
module EX(S: Symantics) = struct open S
  let test1 () = app (lam (fun x -> x)) (bool true)
  let testpowfix () =
       lam (fun x -> fix (fun self -> lam (fun n ->
        if_ (leq n (int 0)) (fun () -> int 1)
            (fun () -> mul x (app self (add n (int (-1))))))))
  let testpowfix7 =  lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
\end{floatrule}
\caption{A simple (Meta)OCaml embedding of our object language}
\label{fig:ocaml-simple}
\end{figure*}
To embed the same object language in (Meta)OCaml, we replace the \ifshort type
class \fi |Symantics| \ifshort\else type class \fi and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until~\S\ref{PE}.  The two differences are:
the additional type parameter |'c|, an
\emph{environment classifier} \cite{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| since OCaml is a call-by-value
language.

@@

Thus we represent an object expression in
OCaml as a functor from |Symantics| to an appropriate semantic domain. This
is essentially the same as the constraint \texttt{Symantics repr =>} in the
Haskell embedding.

@@

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation
\vspace*{-0.8em}\begin{code3}
module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
  let int  (x:int)  = x         let bool (b:bool) = b
  let lam  f        = f         let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee () end
\end{code3}
\vspace*{-0.8em}
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|%
%aplas-specific
: \texttt{module EXR = EX(R)}%
.
%% \begin{code}
%% module EXR = EX(R)
%% \end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
% aplas: commented out
\begin{comment}
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
\end{comment}
%aplas: the following line is for aplas only
It is obvious to the compiler that
%
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
%aplas-specific
(The source code shows a total interpreter |L| that measures
the size of each object term).
%aplas-specific (cont)
We can also generalize from~|R| to all interpreters; the proof
of these propositions follow immediately from the soundness of the
metalanguage's type system.
%end aplas-specific
\begin{proposition}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
% Generalizing from~|R| to all interpreters, we have the following
% broader and more useful, if also more obvious, proposition.
\begin{proposition}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{proposition}
% These propositions follow immediately from the soundness of the
% metalanguage's type system.
% aplas-commented
\begin{comment}
For variety, we show another interpreter, which measures the \emph{size}
of each object term, defined as the number of term
constructors. The following is slightly abbreviated code (see the
accompanying source code for the complete definition).
\begin{code}
module L = struct
  type ('c,'dv) repr = int
  let int (x:int)  = 1
  let lam f        = f 0 + 1
  let app e1 e2    = e1 + e2 + 1
  let fix f        = f 0 + 1
  let mul e1 e2    = e1 + e2 + 1
  let if_ eb et ee = eb + et () + ee () + 1
end
\end{code}
Now the expression
\begin{code}
let module E = EX(L) in E.test1 ()
\end{code}
evaluates to |3|. This interpreter is not only tagless but also
total. It ``evaluates'' even seemingly divergent terms like
\begin{code}
app (fix (fun self -> self)) (int 1)
\end{code}

%aplas \begin{comment}
module EX1(S: Symantics) = struct
 open S
 let tfix () = app (fix (fun self -> self)) (int 1)
end;;
let module E =EX1(R) in E.tfix ();;
let module E =EX1(L) in E.tfix ();;
%%aplas \end{comment}
\end{comment}

\section{A tagless compiler (or, a staged interpreter)}\label{compiler}
\vspace{-5pt}
Besides immediate evaluation, we can compile our object language
into OCaml code using MetaOCaml's staging facilities. MetaOCaml
represents future-stage expressions of type~$t$ at the
present stage as values of type |('c,|$t$|) code| where |'c| is the
environment classifier \cite{WalidPOPL03,calcagno-ml-like}. Code values are created
by a \emph{bracket} form |.<|$e$|>.|, which specifies that the expression~$e$ is to be
evaluated at a future stage. The \emph{escape} |.~|$e$ must occur
within a bracket and specifies that the expression~$e$ must be evaluated
at the current stage; its result, which must be a code value, is
spliced into the code being built by the enclosing bracket. The \emph{run} form |.!|$e$ evaluates
the future-stage code value~$e$ by compiling and linking it at run-time.
The bracket, escape, and run are akin to
quasi-quotation, unquotation, and |eval| of Lisp.

Inserting brackets and escapes appropriately into the
evaluator~|R| above yields the simple compiler~|C| below%
%aplas-specific
, Fig.~\ref{fig:interpreter-C}(a)
.
% fig:interpreter-C part (a) was here
%
\begin{figure}[t]
% this code does not have the 'sv parameter. It shows up later.
(a) \begin{code2}
module C = struct type ('c,'dv) repr = ('c,'dv) code
  let int (x:int)   = .<x>.           let bool (b:bool) = .<b>.
  let lam f         = .<fun x -> .~(f .<x>.)>.
  let app e1 e2     = .<.~e1 .~e2>.
  let fix f =  .<let rec self n = .~(f .<self>.) n in self>.
  let add e1 e2     = .<.~e1 + .~e2>. let mul e1 e2     = .<.~e1 * .~e2>.
  let leq x y       = .<.~x <= .~y>.
  let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>. end
\end{code2}

(b) \begin{code2}
let module E = EX(C) in E.test1 ()
\end{code2}

(c) \begin{code2}
let module E = EX(C) in E.testpowfix7
\end{code2}

(d) \begin{code2}
.<fun x_1 -> (fun x_2 -> let rec self_3 = fun n_4 ->
   (fun x_5 -> if x_5 <= 0 then 1  else x_2 * (self_3 (x_5 + (-1))))
   n_4 in self_3) x_1 7>.
\end{code2}
\caption{The tagless staged interpreter \texttt{C}.}
\label{fig:interpreter-C}
\end{figure}
%
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1|%
%aplas-specific
, Fig.~\ref{fig:interpreter-C}(b),
% fig:interpreter-C part (b) was here
gives the code value |.<(fun x_6 -> x_6) true>.|
of inferred type |('c, bool) C.repr|.  Interpreting |testpowfix7|
with
%aplas-specific
Fig.~\ref{fig:interpreter-C}(c),
% fig:interpreter-C part (c) was here
gives a code value with many apparent $\beta$-redexes
%aplas-specific
Fig.~\ref{fig:interpreter-C}(d).
% fig:interpreter-C part (d) was here
This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| 
% aplas: and does not
% call the interpreter, unlike the recursive calls to |eval0| and
% |eval| in the |L e| lines in \S\ref{tagproblem}.
The resulting code obviously contains no tags and no pattern matching.
%aplas: the line is for aplas. The original paragraph commented out
The accompanying code shows the Haskell implementation. 
\begin{comment}
We have also implemented this compiler in Haskell. 
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|%
\footnote{\texttt{ByteCode} can be mapped to AST of Template Haskell
  (TH). The output of our compiler will then be assuredly type-correct TH.}
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. 
We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|.
The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.
\end{comment}

@@

Roughly, a partial evaluator interprets each object term to either
a static (present-stage) term (using~|R|) or
a dynamic (future-stage) term (using~|C|).  To
distinguish between static and dynamic terms, one might define
%aplas-specific
\texttt{data P0 t = S0 (R t) $\mid$ E0 (C t)}.
We then start to define |P0| as an instance of the |Symantics| class.
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R|) if and only if both operands
are static; otherwise we extract the dynamic terms from the operands and
add them using~|C|.
Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(type |C Bool|) and the second and third arguments are a static term
(type |R a|) and a dynamic term (type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\cite{xi-guarded,WalidPOPL03}.
%aplas-else
\begin{comment}
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrI :: P0 Int -> C Int
abstrI (S0 r) = int (unR r)
abstrI (E0 c) = c

abstrB :: P0 Bool -> C Bool
abstrB (S0 r) = bool (unR r)
abstrB (E0 c) = c
\end{code}
We then start to define |P0| as an instance of the |Symantics| class.
\begin{code}
instance Symantics P0 where
  int  x = S0 (int x)
  bool x = S0 (bool x)
  add (S0 e1) (S0 e2) = S0 (add e1 e2)
  add e1 e2 = E0 (add (abstrI e1) (abstrI e2))
\end{code}
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R|) if and only if both operands
are static; otherwise we extract the dynamic terms from the operands and
add them using~|C|.  Thus the two uses of |add| above refer to
|add| for~|R| and |add| for~|C|.

Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(type |C Bool|) and the second and third arguments are a static term
(type |R a|) and a dynamic term (type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\cite{xi-guarded,WalidPOPL03}.\footnote{We note in passing that, if we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\quad\texttt{fmap
f = app (lift f)}}
\end{comment}

>>>>

A popular way to define and implement a language is to embed it in
another \citep{reynolds-definitional}.  Embedding means to represent
terms and values of the \emph{object language} as terms and values in the
\emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
specific object languages because it supports rapid prototyping and integration
with the host environment \citep{hudak-building}.
If the metalanguage supports \emph{staging}, then
the embedding can compile object programs to the metalanguage and avoid the
overhead of interpreting them on the fly \citep{WalidICFP02}.  A staged
definitional interpreter is thus a promising way to build a domain\hyp specific
language (DSL)\@.

@@

We represent object programs using ordinary functions rather than
data constructors.  These functions comprise the entire interpreter, shown
below.
\ifshort
\begin{code3}
let varZ env    = fst env         let b (bv:bool) env = bv
let varS vp env = vp (snd env)    let lam e env   = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code3}
\else
\begin{code}
let varZ env        = fst env
let varS vp env     = vp (snd env)
let b (bv:bool) env = bv
let lam e env       = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code}
\fi
We now represent our sample term $(\fun{x}x)\True$ as
\ifshort
\texttt{let testf1 = app (lam varZ) (b true)}.
\else
\begin{code}
let testf1 = app (lam varZ) (b true)
\end{code}
\fi
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment\ifshort
, |testf1 ()|, obtaining |true|\fi.
\ifshort\else
\begin{code}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code}
\fi
The result has no tags: the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
open term such as
\ifshort
\texttt{app (lam (varS varZ)) (b true)}
\else
|testf3| below
\fi
gives a type error rather than a run-time error.
\ifshort\else
\begin{code}
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code}
\fi
The type error correctly complains
that the initial environment should be a tuple rather than~|()|.
In other words, the term is open.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        even call-by-name (CBN) CPS, so to isolate the evaluation
        order of the object language from that of the metalanguage.\ifshort
\footnote{Due to serious lack of space, 
we refer the reader to the accompanying code for this.}\fi
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
MetaOCaml) and standard Haskell. The complete code is available at
\url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.

@@

To reiterate, we do \emph{not} propose any new language
feature or technique. We use features already present in
mainstream functional languages---Hindley-Milner
type system with either an inference\hyp preserving module system or
constructor classes, as realized in ML and Haskell 98---to
solve the problem that was stated in the published record as
unsolved and likely unsolvable in ML or Haskell 98
without extensions. The simplicity of our solution and its
use of only mainstream features makes it more practical to build typed,
embedded DSLs.
\begin{comment}
We may claim some contribution about type-preserving CPS. Check
related work section in \cite{Guillemette-Monier-PLPV}, especially
check the work of Shao on type-preserving CPS in Flint. The PLPV paper
in related work shows other tasks, including closure conversion, which
we may tackle in our approach. We may be able to write a
type-preserving, assured compiler, whose properties and assured by HM.
\end{comment}

@@

\begin{figure*}[t]
\begin{floatrule}
\begin{code2}
module type Symantics = sig type ('c, 'dv) repr
  val int : int  -> ('c, int) repr
  val bool: bool -> ('c, bool) repr

@@

  val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val mul : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
  val if_ : ('c, bool) repr
            -> (unit -> 'x) -> (unit -> 'x) -> (('c, 'da) repr as 'x)
end
module EX(S: Symantics) = struct open S
  let test1 () = app (lam (fun x -> x)) (bool true)
  let testpowfix () =
       lam (fun x -> fix (fun self -> lam (fun n ->
        if_ (leq n (int 0)) (fun () -> int 1)
            (fun () -> mul x (app self (add n (int (-1))))))))
  let testpowfix7 = lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code2}
\end{floatrule}
\caption{A simple (Meta)OCaml embedding of our object language}
\label{fig:ocaml-simple}
\end{figure*}
To embed the same object language in (Meta)OCaml, we replace the \ifshort type
class \fi |Symantics| \ifshort\else type class \fi and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until~\S\ref{PE}.  The two differences are:
the additional type parameter |'c|, an
\emph{environment classifier} \cite{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| since OCaml is a call-by-value
language.

@@

Thus, we represent an object expression in
OCaml as a functor from |Symantics| to an appropriate semantic domain. This
is essentially the same as the constraint \texttt{Symantics repr =>} in the
Haskell embedding.

@@

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation.
\ifshort
\vspace*{-0.8em}\begin{code3}
module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
  let int  (x:int)  = x         let bool (b:bool) = b
  let lam  f        = f         let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee () end
\end{code3}
\vspace*{-0.8em}
\else
\begin{code}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)

  let int  (x:int)  = x
  let bool (b:bool) = b
  let lam  f        = f
  let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee ()
end
\end{code}
\fi
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding}
with~|R|\ifshort: \texttt{module EXR = EX(R)}\fi.
\ifshort\else
\begin{code}
module EXR = EX(R)
\end{code}
\fi
Thus, |EXR.test1 ()| evaluates to the untagged boolean value |true|.
\ifshort It \else
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it
\fi is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\ifshort
(The source code shows a total interpreter |L| that measures
the size of each object term.)
\fi
\ifshort
We can also generalize from~|R| to all interpreters;
these propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\begin{proposition}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
\ifshort\else
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\fi
\begin{proposition}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{proposition}
\ifshort\else
These propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\ifshort\else
For variety, we show another interpreter, which measures the \emph{size}
of each object term, defined as the number of term
constructors. The following is slightly abbreviated code (see the
accompanying source code for the complete definition).
\begin{code}
module L = struct
  type ('c,'dv) repr = int
  let int (x:int)  = 1
  let lam f        = f 0 + 1
  let app e1 e2    = e1 + e2 + 1
  let fix f        = f 0 + 1
  let mul e1 e2    = e1 + e2 + 1
  let if_ eb et ee = eb + et () + ee () + 1
end
\end{code}
Now the OCaml expression
|let module E = EX(L) in E.test1 ()|
evaluates to |3|. This interpreter is not only tagless but also
total. It ``evaluates'' even seemingly divergent terms like
|app (fix (fun self -> self)) (int 1)|.
\fi

\begin{comment}
\begin{code}
module EX1(S: Symantics) = struct
 open S
 let tfix () = app (fix (fun self -> self)) (int 1)
end;;
let module E =EX1(R) in E.tfix ();;
let module E =EX1(L) in E.tfix ();;
\end{code}
\end{comment}

\section{A tagless compiler (or, a staged interpreter)}\label{compiler}
Besides immediate evaluation, we can compile our object language
into OCaml code using MetaOCaml's staging facilities. MetaOCaml
represents future-stage expressions of type~$t$ at the
present stage as values of type |('c,|$t$|) code| where |'c| is the
environment classifier \cite{WalidPOPL03,calcagno-ml-like}. Code values are created
by a \emph{bracket} form |.<|$e$|>.|, which specifies that the expression~$e$ is to be
evaluated at a future stage. The \emph{escape} |.~|$e$ must occur
within a bracket and specifies that the expression~$e$ must be evaluated
at the current stage; its result, which must be a code value, is
spliced into the code being built by the enclosing bracket. The \emph{run} form |.!|$e$ evaluates
the future-stage code value~$e$ by compiling and linking it at run-time.
The bracket, escape, and run are akin to
quasi-quotation, unquotation, and |eval| of Lisp.

% this code does not have the 'sv parameter. It shows up later.
\ifshort
\begin{SaveVerbatim}{5a}
module C = struct type ('c,'dv) repr = ('c,'dv) code
  let int (x:int)  = .<x>.            let bool (b:bool) = .<b>.
  let lam f        = .<fun x -> .~(f .<x>.)>.
  let app e1 e2    = .<.~e1 .~e2>.
  let fix f =  .<let rec self n = .~(f .<self>.) n in self>.
  let add e1 e2    = .<.~e1 + .~e2>.  let mul e1 e2 = .<.~e1 * .~e2>.
  let leq x y      = .<.~x <= .~y>.
  let if_ eb et ee = .<if .~eb then .~(et ()) else .~(ee ())>. end
\end{SaveVerbatim}
\else
\begin{SaveVerbatim}{5a}
module C = struct
  type ('c,'dv) repr = ('c,'dv) code

  let int (x:int)   = .<x>.
  let bool (b:bool) = .<b>.
  let lam f         = .<fun x -> .~(f .<x>.)>.
  let app e1 e2     = .<.~e1 .~e2>.
  let fix f         = .<let rec self n = .~(f .<self>.) n in self>.
  let add e1 e2     = .<.~e1 + .~e2>.
  let mul e1 e2     = .<.~e1 * .~e2>.
  let leq x y       = .<.~x <= .~y>.
  let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>.
end
\end{SaveVerbatim}
\fi
\begin{SaveVerbatim}{5b}
let module E = EX(C) in E.test1 ()
\end{SaveVerbatim}
\begin{SaveVerbatim}{5c}
let module E = EX(C) in E.testpowfix7
\end{SaveVerbatim}
\begin{SaveVerbatim}{5d}
.<fun x_1 -> (fun x_2 -> let rec self_3 = fun n_4 ->
   (fun x_5 -> if x_5 <= 0 then 1 else x_2 * self_3 (x_5 + (-1)))
   n_4 in self_3) x_1 7>.
\end{SaveVerbatim}

Inserting brackets and escapes appropriately into the
evaluator~|R| above yields the simple compiler~|C|
\ifshort in Fig.~\ref{fig:interpreter-C}(a).\else below.
\UseVerbatim{5a}
\fi
\ifshort
\begin{figure}[t]
(a) \BUseVerbatim{5a}

(b) \BUseVerbatim{5b}

(c) \BUseVerbatim{5c}

(d) \BUseVerbatim{5d}
\caption{The tagless staged interpreter \texttt{C}.}
\label{fig:interpreter-C}
\end{figure}
\fi
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1| with
\ifshort Fig.~\ref{fig:interpreter-C}(b) \else \UseVerbatim{5b}\fi
gives the code value |.<(fun|~|x_6|~|->|~|x_6)| |true>.|
of inferred type |('c,|~|bool)| |C.repr|.  Interpreting |testpowfix7|
\ifshort\pagebreak[3]\fi
with
\ifshort Fig.~\ref{fig:interpreter-C}(c) \else \UseVerbatim{5c}\fi
gives a code value with many apparent $\beta$- and $\eta$-redexes\ifshort,
Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun|~|x_6|~|->|~|x_6|\ifshort\else\
and does not
call the interpreter, unlike the recursive calls to |eval0| and
|eval| in the |L e| lines in \S\ref{tagproblem}\fi.
The resulting code obviously contains no tags and no pattern matching.
\ifshort
The accompanying code shows the Haskell implementation. 
\else

We have also implemented this compiler in Haskell. 
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it: we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.%
\footnote{\texttt{ByteCode} can be mapped to the abstract syntax of Template
    Haskell. The output of our compiler will then be assuredly type-correct
    Template Haskell.}
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. 
We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|.
The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.
\fi

@@

Roughly, a partial evaluator interprets each object term to either
a static (present-stage) term (using\ifshort\else\ the evaluator\fi~|R|) or
a dynamic (future-stage) term (using\ifshort\else\ the compiler\fi~|C|).  To
distinguish between static and dynamic terms, one might define
\ifshort
\texttt{data P0~t~= S0~(R~t)~}\Verb+|+\texttt{ E0~(C~t)}.
\else
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrI :: P0 Int -> C Int
abstrI (S0 r) = int (unR r)
abstrI (E0 c) = c

abstrB :: P0 Bool -> C Bool
abstrB (S0 r) = bool (unR r)
abstrB (E0 c) = c
\end{code}
\fi
We then start to define |P0| as an instance of the |Symantics| class.
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R|) if and only if both operands
are static; otherwise we extract the dynamic terms from the operands and
add them using~|C|.
\ifshort\else
\begin{code}
instance Symantics P0 where
  int  x = S0 (int x)
  bool x = S0 (bool x)
  add (S0 e1) (S0 e2) = S0 (add e1 e2)
  add e1 e2 = E0 (add (abstrI e1) (abstrI e2))
\end{code}
Thus, the two uses of |add| above refer to
|add| for~|R| and |add| for~|C|.

\fi
Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(type |C Bool|) and the second and third arguments are a static term
(type |R a|) and a dynamic term (type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\cite{xi-guarded,WalidPOPL03}.\ifshort\else
\footnote{If we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\quad\texttt{fmap
f = app (lift f)}}\fi

Our |Symantics| class only includes separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus, compiling polymorphic lift requires intensional type
analysis.  To avoid needing polymorphic lift, we turn to
Asai's technique \cite{asai-binding-time,sumii-hybrid}:
build a dynamic term
alongside every static term.
